Architectures for workflow integration in science gateways

### 2016-04-13 paper.tex
Updates in Section 3

Figure~\ref{fig:notations} shows the graphical notations used to
describe the architectures in the reminder of this
Section. Architectures will be described from their main software
components and interactions. Software components include science
gateway and infrastructure in all architectures, and workflow service,
workflow pool and agent are involved in some architectures. The
workflow engine itself is represented by a specific symbol. Software
interactions among these components are described in the next
Section. Abstract interactions are specific types of interactions that
may be implemented by various different software interactions. Red
color is used to represent the components and interactions that need
to be specifically developed to integrate workflow engines. Tasks
refer to computing tasks that are created by the workflow engine and
executed on the infrastructure. Data represents any type of file, or
database that is involved in the workflow execution and stored on the
infrastructure while the workflow executes. It does not cover workflow
parameters such as strings, numbers, etc.

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by science
  administrator and it results in an interface, for instance a web
  form, where users can enter the parameters of the workflow to be
  executed. The interaction has two aspects: (\texttt{a$_1$}) the
  programs used in the workflow are installed on the infrastructure,
  which may or may not require administrative privileges on the
  infrastructure; (\texttt{a$_2$}) the workflow has to be configured
  in the science gateway so that it becomes available to users. Note
  that integrating a workflow is not the same process as integrating a
  workflow \emph{engine}.
\item Task control: operations to manage tasks on the
  infrastructure, including: authentication, submission, monitoring,
  termination, deletion, etc. When the infrastructure is a grid or a
  cloud, it is for instance implemented using libraries such as SAGA,
  DRMAA, OCCI and similar initiatives \todo{refs needed}.
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer data across the
  infrastructure. 
\item Workflow control: operations to execute a workflow with an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (a.k.a. black box) or
  fine-grained (white box). In a coarse-grained model, workflow
  activities are masked and the user only has a global view of the
  workflow execution. In a fine-grained model, user is exposed to the
  workflow topology, i.e. to the outputs of the individual tasks,
  their statuses and so on. 
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task which submitted a sub-task is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to finish before finishing, and conversely all the
  sub-tasks may be killed if the parent task is killed.
\item Pool-agent: specific to the architecture described
  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
  work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. 
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

See Figure~\ref{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal, a controller
in a Ruby on Rails application, and so on. The workflow engine and the
science gateway usually share a database where application, users and
other resources are stored. In this model, task control and data
control are initiated from the science gateway. This is the model
adopted in the Catania Science Gateway Framework~\cite{Ardizzone2012}
(see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
CIPRES~\cite{miller2010creating} and LONI Pipeline
Environment~\cite{dinov2009efficient}.\todo{Add better justification}

@@

See Figure~\ref{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction that might be
implemented as a web-service call (e.g. RESTful or SOAP), as a
command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
workflow activities, or as a grey box that allows for some interaction
with them.  The workflow engine is responsible for controlling the
tasks on the infrastructure, and performing the required data
transfers to execute them. User data is usually managed through the
science gateway, although it might as well be delivered by the
workflow engine directly to the user.

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that use it~\cite{kacsuk2014science}.

@@

See Figure~\ref{archi:sub-task}. The science gateway is responsible to
execute the tasks on the infrastructure, dealing with heterogeneous
batch managers and meta-schedulers. The workflow engine is a
particular task that can submit sub-tasks to the science gateway. The
workflow engine keeps track of the dependencies between the sub-tasks
but their execution is delegated to the science gateway. The science
gateway may implement mechanisms to deal with task dependencies such
as basic dependency lists as available in most batch
managers. Although it has no global vision of the workflow, it can
keep track of the sub-tasks submitted by a given task, for instance to
be able to cancel them when the task is canceled. The fact that tasks
are submitted through the science gateway does not restrict the range
of possible execution models: pilot jobs, for instance, can still be
used.

\todo{Interactions c1 and c2 are merged in c since this is done
  completely by the science gateway}

This model is implemented in CBRAIN~\cite{SHER-14} where it is used to
integrate the PSOM workflow engine and the FSL toolkit through its
\texttt{fsl\_sub} tool. In particular, the CBRAIN-PSOM integration is
described in~\cite{GLAT-16} and uses an agent computing model (a.k.a pilot
jobs).

>>>>

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software components and interactions. Software components include
science gateway and infrastructure in all architectures, and workflow
service, workflow pool and agent are involved in some
architectures. The workflow engine itself is represented by a specific
symbol. Software interactions among these components are described in
the next Section. Abstract interactions are specific types of
interactions that may be implemented by various different software
interactions. Red color is used to represent the components and
interactions that need to be specifically developed to integrate
workflow engines. Tasks refer to computing tasks that are created by
the workflow engine and executed on the infrastructure. Data
represents any type of file, or database that is involved in the
workflow execution and stored on the infrastructure while the workflow
executes. It does not cover workflow parameters such as strings,
numbers, etc. The reminder of this section describes the different
types of interactions involved in the architectures, and details each
architecture.

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may or may not require administrative
  privileges on the infrastructure; (\texttt{a$_2$}) the workflow has
  to be configured in the science gateway so that it becomes available
  to users. Note that integrating a workflow is not the same process
  as integrating a workflow \emph{engine}.
\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it is for instance implemented using libraries such as
  SAGA, DRMAA, OCCI and similar initiatives \todo{refs needed}.
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer data across the infrastructure. The
  infrastructure might offer various data storage backends with
  heterogeneous interfaces. Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to execute a workflow with an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (a.k.a. black box) or
  fine-grained (white box). In a coarse-grained model, workflow
  activities are masked and the user only has a global view of the
  workflow execution. In a fine-grained model, user is exposed to the
  workflow topology, i.e. to the outputs of the individual tasks,
  their statuses and so on. 
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task which submitted a sub-task is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to finish before finishing, and conversely all the
  sub-tasks may be killed if the parent task is killed.
\item Pool-agent: specific to the architecture described
  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
  work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. 
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

\todo{In the following sections, for each architecture, add a
  description of a real system that implements it: integrated: ?,
  service: VIP, sub-task: CBRAIN-PSOM, pool: SHIWA pool, nested
  workflow: SHIWA CGI, import: SHIWA FGI}

@@

See Figure~\ref{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
CIPRES~\cite{miller2010creating} and LONI Pipeline
Environment~\cite{dinov2009efficient}.

@@

See Figure~\ref{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g. RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
workflow activities, or as a white box that allows for some
interaction with them. The workflow engine is responsible for
controlling the tasks on the infrastructure (\texttt{b}) and for
performing the required data transfers to execute them
(\texttt{c$_2$}). User data is usually managed through the science
gateway (\texttt{c$_1$}), although it might as well be delivered by
the workflow engine directly to the user.

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that rely on
WS-PGRADE/gUSE~\cite{kacsuk2014science}. Figure~\ref{fig:vip-architecture}
shows the architecture of the Virtual Imaging Platform, where the
MOTEUR workflow engine~\cite{GLAT-08i} is invoked as a service at step
\texttt{3} (interaction \texttt{d} on
Figure~\ref{fig:architectures}). Step \texttt{2} on
Figure~\ref{fig:vip-architecture} corresponds to interaction
\texttt{a$_1$}, and step \texttt{4} maps to interactions \texttt{b}
and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
the tasks and embedded in their descriptions).

\begin{figure*}
\def\svgwidth{2\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform that integrates
  the MOTEUR workflow engine through service invocation.}
\label{fig:vip-architecture}
\end{figure*}

@@

See Figure~\ref{archi:sub-task}. The workflow engine is a particular
task that can submit sub-tasks to the science gateway through
interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks but their execution is delegated to
the science gateway that executes them on the infrastructure through
interaction \texttt{b}. Although the science gateway has no global
vision of the workflow, it can keep track of the sub-tasks submitted
by a given task, for instance to be able to cancel them when the task
is canceled. The science gateway may also implement mechanisms to
facilitate the handling of task dependencies, for instance basic
dependency lists as available in most batch managers (see for instance
attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

The science gateway also transfers both user and workflow data across
the infrastructure, so that interactions \texttt{c$_1$} and
\texttt{c$_2$} are both covered by \texttt{c}. In practice, both
\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
pieces of code.

The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
where it is used to integrate the PSOM workflow
engine~\cite{bellec2012pipeline} and the FSL toolkit. The CBRAIN-PSOM
integration~\cite{GLAT-16} is illustrated on
Figure~\ref{fig:cbrain-psom-architecture}. On this Figure, the science
gateway is represented by components \texttt{CBRAIN portal} and
\texttt{CBRAIN execution server} and the infrastructure consists of
\texttt{Storage servers} and \texttt{Computing cluster with shared
  file system}. Interaction \texttt{b} on
Figure~\ref{fig:architectures} is implemented by steps \texttt{4},
\texttt{5}, \texttt{7} and \texttt{8}. Interaction \texttt{c} is
implemented by steps \texttt{3} and \texttt{10}. Interaction
\texttt{e} is implemented by step \texttt{6}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
the a master coordinates workflow execution by submitting workers
and establishing direct communication channels with them. Note how
this peculiar execution model is totally supported by the sub-tasking
architecture.

The CBRAIN-FSL integration is also remarkable as it allows to leverage
FSL pipelines that are written in a very low-level workflow language
(Linux executables and script) that submits tasks uniformly through a
specific tool called \texttt{fsl\_sub}. To integrate FSL in CBRAIN,
\texttt{fsl\_sub} was modified to implement interaction
\texttt{e}\footnote{See \texttt{fsl\_sub} modified script available at
  \url{https://github.com/aces/cbrain-plugins-neuro}} \todo{Add a
  diagram and maybe a code excerpt to illustrate that.}.

\begin{figure}
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-PSOM.pdf_tex}
\caption{Architecture of the CBRAIN-PSOM integration that illustrates
  the sub-tasking model (Figure extracted from~\cite{GLAT-16}). 1:
  User sends data and pipeline execution request to storage server(s)
  and CBRAIN portal. 2: CBRAIN portal sends execution request to
  execution server on cluster. 3: Execution server transfers data from
  storage server(s). 4-5: Execution server starts PSOM master via task
  scheduler. 6: PSOM master submits PSOM workers to execution
  server. 7-8: Execution server starts PSOM workers through task
  scheduler. 9: PSOM master and workers execute pipeline. 10.
  Execution server transfers results to storage server(s). \todo{Improve graphics}}
\label{fig:cbrain-psom-architecture}
\end{figure}
